Changes in Faculty Composition and Stagnating State Support for Higher Education

### 2023-02-22 text/paper.tex
adjust README

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy \\ \vspace{0.1cm}
    Economics Department, Cornell University
}
% Title
\title{Changes in Faculty Composition and Stagnating State Support for Higher Education
    }
\date{9 December 2022%\today
    \thanks{
    I thank Professor Evan Riehl for guidance on this project, Michael Lovenheim, Ronald Ehrenberg, Douglas Miller, Francine Blau, Levon Barseghyan, and Ryan Dycus for various comments on the project's direction, as well as seminar participants at Cornell University (Autumn 2022) for helpful discussion.
    This project's Github repository, with materials and (publicly available) data for replication, is available at 
    \url{https://github.com/shoganhennessy/state-faculty-composition}.
    Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.
    } \\ \vspace{0.1cm}
    \href{https://github.com/shoganhennessy/state-faculty-composition/blob/main/text/paper.pdf}{Newest version available here.}
}
\rhead{9 December 2022}%\today}
\lhead{Faculty Composition and State Support for Higher Education}
% Begin
\begin{document}
\clearpage
\maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
    \input{sections/00-abstract.tex}
\end{abstract}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy \\ \vspace{0.1cm}
    Economics Department, Cornell University
}
% Title
\title{Changes in Faculty Composition and Stagnating State Support for Higher Education
    }
\date{\today\thanks{
    I thank Professor Evan Riehl for guidance on this project, Michael Lovenheim, Ronald Ehrenberg, Douglas Miller, Francine Blau, Levon Barseghyan, and Ryan Dycus for various comments on the project's direction, as well as seminar participants at Cornell University (Autumn 2022) for helpful discussion.
    This project's Github repository, with materials and (publicly available) data for replication, is available at 
    \url{https://github.com/shoganhennessy/state-faculty-composition}.
    Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.
    } \\ \vspace{0.1cm}
    \href{https://github.com/shoganhennessy/state-faculty-composition/blob/main/text/paper.pdf}{Newest version available here.}
}
\rhead{\today}
\lhead{Faculty Composition and State Support for Higher Education.}
% Begin
\begin{document}
\clearpage
\maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
    \input{sections/00-abstract.tex}
\end{abstract}

### 2023-02-22 text/sections/04-empirical.tex
adjust README

\cite{NBERw23736,chakrabarti2018effect,NBERw27885} address endogeneity in public university finances by exploiting a shift-share instrument for changes in state-level funding interacted with university reliance on state funding in a base period.
\begin{align}
    \label{eqn:public-instrument}
    Z_{i,t} &\coloneqq - \log \left[
    \left( \frac{\text{Total State Funding}_{s(i),t}}{\text{Student Population}_{s(i),t}} \right)
    \sum_{\tau = 0}^{3} \frac 14
    \left( \frac{\text{State Funding}_{i,1990 + \tau}}{\text{Total Revenues}_{i,1990 + \tau}} \right) \right]
\end{align}
The system exploits the fact that institutions who rely on state appropriations more will be affected by state appropriation shocks.
$Z_{i,t}$ is the instrument for (log) state appropriations for institution $i$ in year $t$, interacting the average funding for universities in state $s(i)$ with reliance on state funding relative to total revenues, averaged across the base years 1990--1993.\footnote{
    1990--1993 are defined as data for public university finance data are most comparable (i.e. without many missing values) beginning in 1990.
    \cite{NBERw23736} use the single year 1990 as the base year, though I use the four years to ameliorate missing values in the single year of 1990.
    Results are similar in either specification.
}
$Z_{i,t}$ is contructed as negative to reflect the fact that the long term trend in, and most of the short-run shocks to, state funding for higher education has been negative.
State funding has been been falling, so that the instrument describes shocks to university revenues, mostly in a negative direction.

@@

The first-stage is then as follows, including institution and year fixed effects.
\begin{equation}
    \label{eqn:firststage}
    X_{i,t} = \eta_i + \zeta_t + \delta Z_{i,t} + \epsilon_{i,t}
\end{equation}
We note the conditions for exogeneity in the instrument (following the discussion presented by \citealt{NBERw27885}).
The instrument is exogenous if state policy decisions for funding of public universities are uncorrelated with unobserved institutional changes of any specific college or university in the state \citep{borusyak2022quasi}.
This assumption is plausible given that the majority of states have multiple (i.e. more than five) public universities, without any single university campus receiving the majority of state funding within any single state.
Secondly, addressing endogeneity in university revenues by the shift-share identification strategy requires exogeneity in either the base-line share or shift component of the instrument.
In this case, we satisfy the second: universities' institutional-level decisions are not correlated with contemporaneous or upcoming shocks to state appropriations.\footnote{
    It would be plausible to consider the case that universities make institutional-level decisions in a consistently different manner to those 
    with differing reliance on state appropriations in 1990, so that exogeneity by the base-line share is not plausible here.
}

@@

\autoref{tab:firststage-reg} presents results of the first-stage regression, separately with and without a control for tuition revenue per student, plus institution and year fixed effects.\footnote{
    Representations for frequentist significance levels (i.e. 10, 5, 1\% etc.) are omitted here, and in all following tables.
}
We note columns (1) and (2) estimate that a shock to appropriations (per student in the entire state) of 10\% is associated with 9.8\% change in state funding per student at the university; the instrument is strong, and we note similarity in estimates with and without inclusion of fixed effects.
Columns (3) and (4) include the tuition revenue control (explained in \autoref{foot:control}) to exhibit estimates with the inclusion of this possible collider or bad control.
Column (3) shows similar estimates to columns (1), (2) thanks to inclusion of fixed effects, so that the fixed effects was effective in soaking variation in per-student tuition revenues at the institution-year level.
Column (1) represents the estimates for \autoref{eqn:firststage} with fixed effects, omitting the tuition revenue control, and is the preferred form that I proceed with in the following.

@@

\autoref{fig:firststage-lp} presents local projection estimates for the staying-power of the state appropriation shock to university revenues, with each year showing the years following effect of a shock to state reveunes in year zero.
We see a decaying, yet real and positive, effect of the shock on revenues for the next 10 years, illustrating that the state appropriation shock is a strong, if fading, instrument for state funding per student and justifying its use in local projection IV estimation.

@@

The primary empirical model combines the instrumental variable for state appropriation shocks with the empirical model for the effects of state funding at the university-level --- i.e. parameter $\beta$ in the following.\footnote{
    It is important to note the treatment effect isolated here; the instrumental variables approach identifies the local average treatment effects (LATE) specific to the instrument.
    So we interpret this treatment effect as a university's response in employment count and average salaries to state funding changes, changes specific to state appropriation shocks, among the complier group --- i.e. universities with any exposure to shocks, assuming no universities increase employment or salaries in response to negative revenue shocks.
    \cite{mogstad2021causal} provide a full discussion of the LATE with instruments and multiple monotonicity conditions.
}
\begin{eqnarray}
    \label{eqn:secondstage1}
    X_{i,t} &=& \eta_i + \zeta_t + \delta Z_{i,t} + \epsilon_{i,t} \\
    \label{eqn:secondstage2}
    Y_{i,t} &=& \alpha_i + \gamma_t + \beta \widehat X_{i,t} + \varepsilon_{i,t}
\end{eqnarray}
I estimate the system by two stage least squares, including institution and year fixed effects, and investigate outcomes at the university-level to analyse effects on the university as a result of changes in revenues.
Additionally, I estimate the model via local projections \citep{jorda2005,miller2022} to investigate whether the effects of revenues on faculty composition linger for multiple years after the original appropriation shock.
Regarding outcomes, I focus on the composition of the professors employed at the university by analysing (log) count per student, and average salaries paid to, professors employed by the university.

>>>>

\cite{NBERw23736,chakrabarti2018effect,NBERw27885} address endogeneity in public university finances by exploiting a shift-share instrument for changes in state-level funding interacted with university reliance on state funding in a base period.
\begin{align}
    \label{eqn:public-instrument}
    Z_{i,t} &\coloneqq - \log \left[
    \left( \frac{\text{Total State Funding}_{s(i),t}}{\text{Student Population}_{s(i),t}} \right)
    \sum_{\tau = 0}^{3} \frac 14
    \left( \frac{\text{State Funding}_{i,1990 + \tau}}{\text{Total Revenues}_{i,1990 + \tau}} \right) \right]
\end{align}
The system exploits the fact that institutions who rely on state appropriations more will be affected by state appropriation shocks.
$Z_{i,t}$ is the instrument for (log) state appropriations for institution $i$ in year $t$, interacting the average funding for universities in state $s(i)$ with reliance on state funding relative to total revenues, averaged across the base years 1990--1993.\footnote{
    1990--1993 are defined as data for public university finance data are most comparable (i.e. without many missing values) beginning in 1990.
    \cite{NBERw23736} use the single year 1990 as the base year, though I use the four years to ameliorate missing values in the single year of 1990.
    Results are similar in either specification.
}
$Z_{i,t}$ is constructed as negative to reflect the fact that the long term trend in, and most of the short-run shocks to, state funding for higher education has been negative.
State funding has been been falling, so that the instrument describes shocks to university revenues, mostly in a negative direction.

@@

The first-stage is then as follows, including institution and year fixed effects.
\begin{equation}
    \label{eqn:firststage}
    X_{i,t} = \eta_i + \zeta_t + \delta Z_{i,t} + \epsilon_{i,t}
\end{equation}
We note the conditions for exogeneity in the instrument (following the discussion presented by \citealt{NBERw27885}).
The instrument is exogenous if state policy decisions for funding of public universities are uncorrelated with unobserved institutional changes of any specific college or university in the state \citep{borusyak2022quasi}.
This assumption is plausible given that the majority of states have multiple (i.e., more than five) public universities, without any single university campus receiving the majority of state funding within any single state.
Secondly, addressing endogeneity in university revenues by the shift-share identification strategy requires exogeneity in either the base-line share or shift component of the instrument.
In this case, we satisfy the second: universities' institutional-level decisions are not correlated with contemporaneous or upcoming shocks to state appropriations.\footnote{
    It would be plausible to consider the case that universities make institutional-level decisions in a consistently different manner to those 
    with differing reliance on state appropriations in 1990, so that exogeneity by the base-line share is not plausible here.
}

@@

\autoref{tab:firststage-reg} presents results of the first-stage regression, separately with and without a control for tuition revenue per student, plus institution and year fixed effects.\footnote{
    Representations for frequentist significance levels (i.e. 10, 5, 1\% etc.) are omitted here, and in all following tables.
}
We note columns (1) and (2) estimate that a shock to appropriations (per student in the entire state) of 10\% is associated with 9.8\% change in state funding per student at the university; the instrument is strong, and we note similarity in estimates with and without inclusion of fixed effects.
Columns (3) and (4) include the tuition revenue control (explained in \autoref{foot:control}) to exhibit estimates with the inclusion of this possible collider or bad control.
Column (3) shows similar estimates to columns (1), (2) thanks to inclusion of fixed effects, so that the fixed effects was effective in soaking variation in per-student tuition revenues at the institution-year level.
Column (1) represents the estimates for \autoref{eqn:firststage} with fixed effects, omitting the tuition revenue control, and is the preferred form that I proceed with.

@@

\autoref{fig:firststage-lp} presents local projection estimates for the staying-power of the state appropriation shock to university revenues, with each year showing the years following effect of a shock to state revenues in year zero.
We see a decaying, yet real and positive, effect of the shock on revenues for the next 10 years, illustrating that the state appropriation shock is a strong, if fading, instrument for state funding per student and justifying its use in local projection IV estimation.

@@

The primary empirical model combines the instrumental variable for state appropriation shocks with the empirical model for the effects of state funding at the university-level --- i.e., parameter $\beta$ in the following.\footnote{
    It is important to note the treatment effect isolated here; the instrumental variables approach identifies the local average treatment effects (LATE) specific to the instrument.
    So we interpret this treatment effect as a university's response in employment count and average salaries to state funding changes, changes specific to state appropriation shocks, among the complier group --- i.e., universities with any exposure to shocks, assuming no universities increase employment or salaries in response to negative revenue shocks.
    \cite{mogstad2021causal} provide a full discussion of the LATE with instruments and multiple monotonicity conditions.
}
\begin{eqnarray}
    \label{eqn:secondstage1}
    X_{i,t} &=& \eta_i + \zeta_t + \delta Z_{i,t} + \epsilon_{i,t} \\
    \label{eqn:secondstage2}
    Y_{i,t} &=& \alpha_i + \gamma_t + \beta \widehat X_{i,t} + \varepsilon_{i,t}
\end{eqnarray}
I estimate the system by two stage least squares, including institution and year fixed effects, and investigate outcomes at the university-level to analyse effects on the university as a result of changes in revenues.
Additionally, I estimate the model via local projections \citep{jorda2005,miller2022} to investigate whether the effects of revenues on faculty composition linger for multiple years after the original appropriation shock.
Regarding outcomes, I focus on the composition of the professors employed at the university by analysing (log) count per student, and average salaries paid to, professors employed by the university.

### 2023-02-22 text/sections/06-discussion.tex
adjust README

@@

These examples show that Illinois public universities were effected by the negative shock to state funding for higher education, and instituted policy changes aimed at affecting faculty outcomes investigated in \autoref{sec:results}.
Yet the only measurable effect was on count of professor per student (implied via slower hiring), and first-year lecturer salaries.
To investigate these trends for the entire state, \autoref{fig:newhire-illinois} shows the count of professors hired in the years 2011-2021, and their mean salary, by position.
In particular, no position had sustained rises in hiring while enrolment within the university was rising (so that new hires per student necessarily fell), and hiring of new administrators jumped once the Illinois budget crisis ended in 2017.
At the same time, professor new hires' starting salaries remained relatively constant for lecturer and assistant professors, yet rose and fell over the decade for full professors and administrators.

>>>>

\subsection{Faculty Hiring at all US Universities}

Write here about the faculty hiring network data, as provided by \cite{wapman2022quantifying}.

Give a visualisation of the correlation between state funding per student (OLS) and faculty hiring.
Also show the figure of faculty hiring in public vs private: did private unis hire more (in total and per student) over this time period?

Cite Jackson (2021), effects of spending shock.

IV model, where faculty hiring (per student) is the outcome, are presented in \autoref{tab:hiring-shock-reg}.\footnote{
    Give extra details on this specification.
}

\begin{table}[h!]
    \singlespacing
    \centering
    \caption{OLS and 2SLS Estimates for University Faculty Hiring, Total for 2011--2020.}
    \makebox[\textwidth][c]{\input{tables/hiring-shock-reg-fte.tex}}
    \begin{flushleft}
        \footnotesize
        \textbf{Note}: Standard errors are clustered at the state level.
    \end{flushleft}
    \label{tab:hiring-shock-reg}
\end{table}

Note that yearly variation is not observed here, so that only aggregate level, for 180 universities, can be considered.
Leave it to further research to delve deeper into the role that hiring plays as the primary causal mediator for the effect of state funding shocks on faculty composition.

\subsection{Faculty Hiring at Illinois Public Universities}

@@

These examples show that Illinois public universities were affected by the negative shock to state funding for higher education, and instituted policy changes aimed at affecting faculty outcomes investigated in \autoref{sec:results}.
Yet the only measurable effect was on count of professor per student (implied via slower hiring), and first-year lecturer salaries.
To investigate these trends for the entire state, \autoref{fig:newhire-illinois} shows the count of professors hired in the years 2011-2021, and their mean salary, by position.
In particular, no position had sustained rises in hiring while enrolment within the university was rising (so that new hires per student necessarily fell), and hiring of new administrators jumped once the Illinois budget crisis ended in 2017.
At the same time, professor new hires' starting salaries remained relatively constant for lecturer and assistant professors, yet rose and fell over the decade for full professors and administrators.
